Interspecies comparisons of Mg/Ca ratios in limpet shells

### 2024-06-24 Manuscript.qmd
added iso

---
title: "Interspecies comparisons of Mg/Ca ratios in limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Francisco Zangrando
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Angelica Tivoli
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Ivan Briz-Godino
    affiliations:
      - name: 
        address: 
        city: 
        country: Spain
        postal-code: 
abstract: |
  This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures. While *Patella vulgata* along the Spanish shoreline has since then repeatedly produced reliable correlations between sea surface temperature and Mg/Ca ratios, this is not the case for other patelloid species. *Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results. In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens. Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between δ^18^O and Mg/Ca ratios. By reassessing these previous specimens and including a wider range of  modern and archaeological samples from three patelloid species (*P. vulgata*, *N. deaureata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change and serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using δ^18^O or clumped oxygen isotope analysis.
keywords: 
  - Sclerochronology
  - Limpets
  - Elemental Ratio
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format: html

  # elsevier-pdf:
  #   keep-tex: true
  #   journal:
  #     name: Palaeo^3^
  #     formatting: preprint
  #     model: 3p
  #     cite-style: authoryear
output:
  units: cm
---

@@

D18O <- read_csv(here("data","D18O.csv")) %>% 
  mutate(period=case_when(startsWith(sampleID,"Q") ~ "Archaeological",TRUE ~ "Modern"))

@@

Nine additional shells of *N. magellanica* (n=6) and *N*. *deaureata* (n=3) have been selected for LIBS analysis only, to provide a comparative dataset to the modern *Nacella* sp. shells. These have not been analysed for δ^18^O-values, but rather as a test to see whether patterns visible in the modern Mg/Ca ratio distributions are also preserved and apparent in the archaeological shells. They derive from the site of Heshkaia 35

@@

@@

High-quality oxygen isotope values (δ^18^O) were acquired from previous publications on *P. vulgata* [@Surge2012-ba; @Graniero2017-io], *N. deaureata* and *N. magellanica* [@Nicastro2020-ih], where detailed descriptions of the stable isotope analysis methodology can be found. In short, all shells were sectioned along the main growth axis to expose the internal growth structure. From the exposed section carbonate powder samples were acquired using a micromilling approach with resolutions of 100–200 µm and targeting the calcitic M+2 layer of the shells. Analytical precision of the isotope values was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.

Mg/Ca elemental imaging was carried out using Laser Induced Breakdown Spectroscopy at the Leibniz Zentrum für Archäologie (Mainz-Germany) using previously published methods [@Hausmann2023-ih]. These involve the ablation of the shell section using an infrared (1064 nm) laser pulse (1–2 mJ, 100 Hz) onto a surface area of 20–30 µm to generate a plasma plume. This plume emits light, which is measured using a synchronised spectrometer. The resulting light spectrum quantifies the emission lines of magnesium () and calcium () to determine their intensity ratio. While these two peaks alone do not represent the molar concentrations of both elements (as opposed to calibration free LIBS; e.g. [@Martinez-Minchero2022-jz]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa].

### Modern Specimens

```{r Modern iso ranges}
# Ranges <- D18O %>% 
#   filter(!(sampleID == "ORK_LT5" & d18O < 0.2)) %>% #excluding outlier
# group_by(sampleID,period) %>% 
#   mutate(range=max(d18O)-min(d18O),
#          range=round(range,2)) %>% 
#   select(period, sampleID,range) %>% unique()
# 
# Modern_Ranges <- Ranges %>% 
#   filter(period=="Modern") %>% 
#   arrange(range)

#Interestingly, the overall annual range in δ18O-values on Orkney (~r Modern_Ranges$range[3]‰, excluding the anomalous value of 3.38‰) is comparable to the range in the Nacella sp. shells (r min(Modern_Ranges$range)‰ to r max(Modern_Ranges$range)‰) despite the different SST ranges.

@@

All modern shells show multi-year quasi-sinusoidal records of SST change (@fig-Mod_iso), with *Nacella* sp. shells living 2–3 years and *Patella v*. over 5 years. The δ^18^O values of the Orcadian shell have not been directly compared to the local SST, but accurately reflect the seasonal changes expected barring one extreme, anomalous value in the second summer of recorded growth.

```{r Modern Isotope plot}
#| fig-cap: Isotope values of modern limpet shells from Cambaceres Bay (*a-f*) and Westray (*g*).
#| fig-height: 10
#| fig-width: 10
#| label: fig-Mod_iso
# Filter data for Modern period
modern_data <- D18O %>% filter(period == "Modern")

# Get unique sample IDs
sample_ids <- unique(modern_data$sampleID)

@@

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- modern_data %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_point() +
    xlim(0,15)+
    labs(
      x = "Distance from margin in mm",
      y = expression(delta^18*O~("\u2030")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))

@@

### Archaeological Specimens

```{r Archaeological Isotope plot}
#| fig-cap: Isotope values of Archaeological limpet shells from Quoygrew
#| fig-height: 10
#| fig-width: 10
#| label: fig-Arch_iso
# Filter data for Modern period
arch_data <- D18O %>% filter(period == "Archaeological")

# Get unique sample IDs
sample_ids <- unique(arch_data$sampleID)

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- arch_data %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_point() +
    # xlim(0,20)+
    labs(
      x = "Distance from margin in mm",
      y = expression(delta^18*O~("\u2030")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))

## Patella vulgata

## Nacella deaureata

## Nacella magellanica

@@

>>>>

---
title: "Interspecies comparisons of Mg/Ca ratios in limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Francisco Zangrando
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Angelica Tivoli
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Ivan Briz-Godino
    affiliations:
      - name: 
        address: 
        city: 
        country: Spain
        postal-code: 
abstract: |
  This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures. While *Patella vulgata* along the Spanish shoreline has since then repeatedly produced reliable correlations between sea surface temperature and Mg/Ca ratios, this is not the case for other patelloid species. *Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results. In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens. Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between δ^18^O and Mg/Ca ratios. By reassessing these previous specimens and including a wider range of  modern and archaeological samples from three patelloid species (*P. vulgata*, *N. deaureata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change and serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using δ^18^O or clumped oxygen isotope analysis.
keywords: 
  - Sclerochronology
  - Limpets
  - Elemental Ratio
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Palaeo^3^
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm
---

@@

D18O <- read_csv(here("data","D18O.csv")) %>% 
  mutate(period=case_when(startsWith(sampleID,"Q") ~ "Archaeological",TRUE ~ "Modern")) %>% 
  mutate(species = case_when(
      startsWith(sampleID, "ND") ~ "Nacella deaureata",
      startsWith(sampleID, "NM") ~ "Nacella magellanica",
      TRUE ~ "Patella vulgata"
    ))

@@

Nine additional shells of *N. magellanica* (n=6) and *N*. *deaureata* (n=3) have been selected for LIBS analysis only, to provide a comparative dataset to the modern *Nacella* sp. shells. These have not been analysed for δ^18^O-values, but rather as a test to see whether patterns visible in the modern Mg/Ca ratio distributions are also preserved and apparent in the archaeological shells. They derive from the site of Heshkaia 35 situated within Moat Bay, around 30 km east of where the modern specimens in our study were collected [@Zangrando2014-qi; @Zangrando2021-qg]. With radiocarbon dates putting the shells into the Late Holocene (1300 – 1450 CE) they are of similar age — if somewhat younger — as the assemblage from Quoygrew.

@@

Mg/Ca elemental imaging was carried out using Laser Induced Breakdown Spectroscopy at the Leibniz Zentrum für Archäologie (Mainz-Germany) using previously published methods [@Hausmann2023-ih]. These involve the ablation of the shell section using an infrared (1064 nm) laser (1–2 mJ, 100 Hz) onto a surface area of 20–30 µm to generate a plasma plume. This plume emits light, which is measured using a synchronised spectrometer. The resulting light spectrum quantifies the emission lines of magnesium (MgII; 279.553 nm) and calcium (CaII; 315.887 nm) to determine their intensity ratio. While these two peaks alone do not represent the molar concentrations of both elements (as opposed to calibration free LIBS; e.g. [@Martinez-Minchero2022-jz]), the intensity ratio is linearly correlated to the molar concentration [@Hausmann2017-oa] and can be used as a reliable indicator of Mg/Ca changes within the shell carbonate.

@@

High-quality oxygen isotope values (δ^18^O) were acquired from previous publications on *P. vulgata* [@Surge2012-ba; @Graniero2017-io], *N. deaureata* and *N. magellanica* [@Nicastro2020-ih], in which detailed descriptions of the stable isotope analysis methodology can be found. In short, all shells were sectioned along the main growth axis to expose the internal growth structure. From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic M+2 layer of the shells. Analytical precision of the isotope values was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.

@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential δ^18^O-values of limpet shells by species. Due to the high sampling resolution, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change , with *Nacella* sp. shells living 2–3 years and *Patella v*. regularly over 3 years. In those instances where longer lifetimes are encountered (i.e. specimen QG1-7246-1, @fig-Pat_iso b). Interestingly, the ranges in δ^18^O-values are not similar between the coeval *Nacella* sp. specimens with δ^18^O values for the previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for winter values between 3.0‰ and 2.0‰. *P. vulgata* has similarly variable δ^18^O-values for summer and winter seasons, alas the lack of more that one modern specimens, means that we cannot quantify the variability of recorded annual ranges between specimens.

```{r Nacella_Isotope_plot}
#| fig-cap: Isotope values of Nacella sp. limpet shells from Cambaceres Bay (*a-f*).
#| fig-height: 10
#| fig-width: 10
#| label: fig-Nac_iso
# Filter data for Modern period
Nacella_iso <- D18O %>% filter(grepl("Nacella", species))

# Get unique sample IDs
sample_ids <- unique(Nacella_iso$sampleID)

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- Nacella_iso %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_point() +    labs(
      x = "Distance from margin in mm",
      y = expression(delta^18*O~("\u2030")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))

# Combine plots using patchwork and add letters
wrap_plots(plots,nrow=3) + plot_annotation(tag_levels = 'a')

@@

```{r Patella_Isotope_plot}
#| fig-cap: Isotope values of Patella vulagata limpet shells from Quoygrew (QG, *a-g*) and Rack Wick Bay (ORK, *h*) on Westray, Orkney.
#| fig-height: 10
#| fig-width: 10
#| label: fig-Pat_iso
# Filter data for Modern period
Patella_iso <- D18O %>% filter(grepl("Patella", species))

# Get unique sample IDs
sample_ids <- unique(Patella_iso$sampleID)

@@

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- Patella_iso %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_point() +
    # xlim(0,20)+
    labs(
      x = "Distance from margin in mm",
      y = expression(delta^18*O~("\u2030")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))

@@

## *Patella vulgata*

## *Nacella deaureata*

## *Nacella magellanica*

@@

# References
```{r Modern iso ranges}
# Ranges <- D18O %>%
#   filter(!(sampleID == "ORK_LT5" & d18O < 0.2)) %>% #excluding outlier
# group_by(sampleID,period) %>%
#   mutate(max=max(d18O),
#          min=min(d18O)) %>%
#   select(min, sampleID,max) %>% unique()
# 
# Modern_Ranges <- Ranges %>%
#   filter(period=="Modern") %>%
#   arrange(range)

# Interestingly, the overall annual range in δ^18^O-values on Orkney (\~`r Modern_Ranges$range[3]`‰, excluding the anomalous value of 3.38‰) is comparable to the range in the Nacella sp. shells (`r min(Modern_Ranges$range)`‰ to `r max(Modern_Ranges$range)`‰) despite the different SST ranges.

### 2024-06-24 Manuscript.tex
added iso

Problematically, links between Mg/Ca ratios and SST changes in many
marine mollusc shells, including limpets, have shown to be unreliable
\citep{Surge2008-ri, Wanamaker2008-zl, Schone2010-yl, Freitas2012-tx, Graniero2015-zv, Poulain2015-dg, Vihtakari2017-wd}.
This is particularly the case where there is little available additional
information on metabolic processes, organic components of the shell
matrix, intra-increment and intra-shell variability, and growth rates,
which can independently and unpredictably affect Mg/Ca ratios and
confound their interpretation as temperature proxy. Confoundingly,
multiple different temperature equations have been found for the same
species (see {[}\citet{Freitas2012-tx}; @Vihtakari2017-wd{]} and
references therein). Coeval specimens sharing one locality can also show
differences in their relation to SST \citep{Hausmann2019-fi}. Where the
use of Mg/Ca as a palaeotemperature proxy was successful, anomalous
patterns in some specimens had still to be filtered out by hand,
reducing the overall robustness of the results of those successful
studies \citep{Ferguson2011-zl}.

While recent research particularly of \emph{Patella} sp. in the
Mediterranean and Southwest Europe have provided promising results
\citep{Hausmann2019-fi, Garcia-Escarzaga2015-jc, Garcia-Escarzaga2018-nf}.
There remains unclarity for Atlantic limpet species, particularly since
it was last shown here, that they are not reliable recorders of
palaeotemperature \citep{Graniero2015-zv}.

@@

Relying and adding onto published datasets, allowed us to simultaneously
avoid costs for new high-resolution δ\textsuperscript{18}O-data, use
real-world examples, and also provide pilot-data for areas of existing
research interest. Generally establishing the usefulness and reliabilty
of Mg/Ca as SST proxy in limpets should help to provide a platform for
future research and an important stepping stone to better understand
elemental ratios in other marine mollusc shells.

@@

In total, we studied 16 archaeological specimens. Of those, seven come
from Quoygrew on Westray dating to 900 -- 1200 CE. These shells have
previously been studied to characterise seasonal temperature change
during the Medieval Climate Anomaly and the Little Ice Age
\citep{Surge2012-ba}. Together they show a range of lifespans, with some
comparatively long for their size (QG1-7246-1, 12 years within 16 mm of
growth) making them particularly susceptible to time-averaging. That
said, the majority are under 5 years old. These shells are an ideal
comparative dataset from an archaeological assemblage to compare to the
modern shell ORK\_LT5 \citep{Graniero2017-io}.

Nine additional shells have been selected for LIBS analysis only, to
provide a comparative dataset to the modern \emph{Nacella} sp. shells.
These have not been analysed for δ\textsuperscript{18}O-values, but
rather as a test to see whether patterns visible in the modern Mg/Ca
ratio distributions are also preserved and apparent in the
archaeological shells.

@@

@@

High-quality oxygen isotope values (δ\textsuperscript{18}O) were
acquired from previous publications on \emph{P. vulgata}
\citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaureata} and \emph{N.
magellanica} \citep{Nicastro2020-ih}, where detailed descriptions of the
stable isotope analysis methodology can be found. In short, all shells
were sectioned along the main growth axis to expose the internal growth
structure. From the exposed section carbonate powder samples were
acquired using a micromilling approach with resolutions of 100--200 µm
and targeting the calcitic M+2 layer of the shells. Analytical precision
of the isotope values was consistently at 0.05--0.10‰ and values are
reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee
Belemnite) standard.

Mg/Ca elemental imaging was carried out using Laser Induced Breakdown
Spectroscopy at the Leibniz Zentrum für Archäologie (Mainz-Germany)
using previously published methods \citep{Hausmann2023-ih}. These
involve the ablation of the shell section using an infrared (1064 nm)
laser pulse (1--2 mJ, 100 Hz) onto a surface area of 20--30 µm to
generate a plasma plume. This plume emits light, which is measured using
a synchronised spectrometer. The resulting light spectrum quantifies the
emission lines of magnesium () and calcium () to determine their
intensity ratio. While these two peaks alone do not represent the molar
concentrations of both elements (as opposed to calibration free LIBS;
e.g. \citep{Martinez-Minchero2022-jz}), the intensity ratio is linearly
correlated to the molar concentration \citep{Hausmann2017-oa}.

\subsubsection{Modern Specimens}\label{modern-specimens-1}

All modern shells show multi-year quasi-sinusoidal records of SST change
(Figure~\ref{fig-Mod_iso}), with \emph{Nacella} sp. shells living 2--3
years and \emph{Patella v}. over 5 years. The δ\textsuperscript{18}O
values of the Orcadian shell have not been directly compared to the
local SST, but accurately reflect the seasonal changes expected barring
one extreme, anomalous value in the second summer of recorded growth.

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Mod_iso-1.pdf}

@@

\caption{\label{fig-Mod_iso}Isotope values of modern limpet shells from
Cambaceres Bay (\emph{a-f}) and Westray (\emph{g}).}

@@

\subsubsection{Archaeological
Specimens}\label{archaeological-specimens-1}

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Arch_iso-1.pdf}

@@

\caption{\label{fig-Arch_iso}Isotope values of Archaeological limpet
shells from Quoygrew}

@@

\subsection{Patella vulgata}\label{patella-vulgata}

\subsection{Nacella deaureata}\label{nacella-deaureata}

\subsection{Nacella magellanica}\label{nacella-magellanica}

@@

  \bibliography{bibliography.bib}

>>>>

Problematically, links between Mg/Ca ratios and SST changes in many
marine mollusc shells, including limpets, have shown to be unreliable
\citep{Surge2008-ri, Wanamaker2008-zl, Schone2010-yl, Freitas2012-tx, Graniero2015-zv, Poulain2015-dg, Vihtakari2017-wd}.
This is particularly the case where there is little available additional
information on metabolic processes, organic components of the shell
matrix, intra-increment and intra-shell variability, and growth rates,
which can independently and unpredictably affect Mg/Ca ratios and
confound their interpretation as temperature proxy. Confusingly,
multiple different temperature equations have been found for the same
species (see \citep{Freitas2012-tx, Vihtakari2017-wd} and references
therein). Coeval specimens sharing one locality can also show
differences in their relation to SST \citep{Hausmann2019-fi}. Where the
use of Mg/Ca as a palaeotemperature proxy was successful, anomalous
patterns in some specimens had still to be filtered out by hand,
reducing the overall robustness of the results of those successful
studies \citep{Ferguson2011-zl}.

While recent research particularly of \emph{Patella} sp. in the
Mediterranean and Southwest Europe have provided promising results
\citep{Hausmann2019-fi, Garcia-Escarzaga2015-jc, Garcia-Escarzaga2018-nf}.
There remains a lack of clarity for Atlantic limpet species,
particularly since it was last shown here, that they are not reliable
recorders of palaeotemperature \citep{Graniero2015-zv}.

@@

By relying and adding onto published datasets, we were able to
simultaneously avoid costs for new high-resolution
δ\textsuperscript{18}O-data, to use real-world examples, and to also
provide pilot-data for areas of existing research interest. Generally
establishing the usefulness and reliability of Mg/Ca as SST proxy in
limpets should help to provide a platform for future research and an
important stepping stone to better understand elemental ratios in other
marine mollusc shells.

@@

In total, we studied 16 archaeological specimens. Of those, seven
\emph{P. vulgata} come from Quoygrew on Westray dating to 900 -- 1200
CE. These shells have previously been studied to characterise seasonal
temperature change during the Medieval Climate Anomaly and the Little
Ice Age \citep{Surge2012-ba}. Together they show a range of lifespans,
with some comparatively long for their size (QG1-7246-1, 12 years within
16 mm of growth) making them particularly susceptible to time-averaging.
That said, the majority are under 5 years old. These shells are an ideal
comparative dataset from an archaeological assemblage to compare to the
modern shell ORK\_LT5 \citep{Graniero2017-io}.

Nine additional shells of \emph{N. magellanica} (n=6) and \emph{N}.
\emph{deaureata} (n=3) have been selected for LIBS analysis only, to
provide a comparative dataset to the modern \emph{Nacella} sp. shells.
These have not been analysed for δ\textsuperscript{18}O-values, but
rather as a test to see whether patterns visible in the modern Mg/Ca
ratio distributions are also preserved and apparent in the
archaeological shells. They derive from the site of Heshkaia 35 situated
within Moat Bay, around 30 km east of where the modern specimens in our
study were collected \citep{Zangrando2014-qi, Zangrando2021-qg}. With
radiocarbon dates putting the shells into the Late Holocene (1300 --
1450 CE) they are of similar age --- if somewhat younger --- as the
assemblage from Quoygrew.

@@

Mg/Ca elemental imaging was carried out using Laser Induced Breakdown
Spectroscopy at the Leibniz Zentrum für Archäologie (Mainz-Germany)
using previously published methods \citep{Hausmann2023-ih}. These
involve the ablation of the shell section using an infrared (1064 nm)
laser (1--2 mJ, 100 Hz) onto a surface area of 20--30 µm to generate a
plasma plume. This plume emits light, which is measured using a
synchronised spectrometer. The resulting light spectrum quantifies the
emission lines of magnesium (MgII; 279.553 nm) and calcium (CaII;
315.887 nm) to determine their intensity ratio. While these two peaks
alone do not represent the molar concentrations of both elements (as
opposed to calibration free LIBS; e.g.
\citep{Martinez-Minchero2022-jz}), the intensity ratio is linearly
correlated to the molar concentration \citep{Hausmann2017-oa} and can be
used as a reliable indicator of Mg/Ca changes within the shell
carbonate.

@@

High-quality oxygen isotope values (δ\textsuperscript{18}O) were
acquired from previous publications on \emph{P. vulgata}
\citep{Surge2012-ba, Graniero2017-io}, \emph{N. deaureata} and \emph{N.
magellanica} \citep{Nicastro2020-ih}, in which detailed descriptions of
the stable isotope analysis methodology can be found. In short, all
shells were sectioned along the main growth axis to expose the internal
growth structure. From the exposed section carbonate powder samples were
acquired using a micromill with resolutions of 100--200 µm and by
targeting the calcitic M+2 layer of the shells. Analytical precision of
the isotope values was consistently at 0.05--0.10‰ and values are
reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee
Belemnite) standard.

Figure~\ref{fig-Nac_iso} and Figure~\ref{fig-Pat_iso} show the
previously acquired sequential δ\textsuperscript{18}O-values of limpet
shells by species. Due to the high sampling resolution, all shells show
seasonally resolved and multi-year quasi-sinusoidal records of SST
change , with \emph{Nacella} sp. shells living 2--3 years and
\emph{Patella v}. regularly over 3 years. In those instances where
longer lifetimes are encountered (i.e.~specimen QG1-7246-1,
Figure~\ref{fig-Pat_iso} b). Interestingly, the ranges in
δ\textsuperscript{18}O-values are not similar between the coeval
\emph{Nacella} sp. specimens with δ\textsuperscript{18}O values for the
previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for
winter values between 3.0‰ and 2.0‰. \emph{P. vulgata} has similarly
variable δ\textsuperscript{18}O-values for summer and winter seasons,
alas the lack of more that one modern specimens, means that we cannot
quantify the variability of recorded annual ranges between specimens.

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Nac_iso-1.pdf}

@@

\caption{\label{fig-Nac_iso}Isotope values of Nacella sp. limpet shells
from Cambaceres Bay (\emph{a-f}).}

@@

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Pat_iso-1.pdf}

@@

\caption{\label{fig-Pat_iso}Isotope values of Patella vulagata limpet
shells from Quoygrew (QG, \emph{a-g}) and Rack Wick Bay (ORK, \emph{h})
on Westray, Orkney.}

@@

\subsection{\texorpdfstring{\emph{Patella
vulgata}}{Patella vulgata}}\label{patella-vulgata}

\subsection{\texorpdfstring{\emph{Nacella
deaureata}}{Nacella deaureata}}\label{nacella-deaureata}

\subsection{\texorpdfstring{\emph{Nacella
magellanica}}{Nacella magellanica}}\label{nacella-magellanica}

@@

\renewcommand\refname{References}
  \bibliography{bibliography.bib}
